MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments

### 2022-03-16 main/MSstatsPTM_Main.tex
Datasets

\todo{This section would benefit from more biology/technology background, and more extensive references}
Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of bimolecular events. Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen}. Localizing the modification sites, or characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions, provides important insights~\cite{Mann}. For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}. Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale \todo{would be good to have a reference}.

Bottom-up liquid chromatography coupled with tandem mass spectrometry (LC-MS/MS) is a tool-of-choice for unbiased and large-scale identification and quantification of proteins and their PTMs~\cite{Kall:2011ub,Roepstorff}. However, LC-MS-based interrogation of the modified proteome is challenging, for a number of reasons. First, the relatively lower abundance of modified proteoforms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads \cite{Huang:2014}. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities. Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide) \cite{Mann}. Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complication the interpretation of the results \cite{Olsen:2013}. Finally, technological aspects of bottom-up MS experiments, and presence of labeling, e.g. by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.

The technological challenges in PTM identification and quantification translate themselves into challenges of statistical data analyses. Yet frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. In particular, methods such as t-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~ \cite{Ritchie_15a}. Such methods do not fully account for all the sources of uncertainty and variation in the experiment. As the result, they are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$ \cite{Choi:2014,Huang:2020}.

@@

The benchmark datasets in this manuscript provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods. Two computer simulations varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies. Table \ref{tab:dataDescription} summarizes the experiments. Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}. Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).

%%%% Summary of controlled mixtures
\begin{table}[h!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of Bio. &  Data \\
&  & Design & Conditions & Replicates & availability \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset in this manuscript.
``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. 
}
\label{tab:dataDescription}
\end{table}

@@

The simulation mimicked a label-free experiment. Synthetic datasets had a high number of modified peptide features and no missing values, while varying numbers of replicates and conditions. Specifically, each dataset had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs. 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions. Of the 500 differential PTMs, the fold change of 250 were fully masked by changes in the unmodified protein. Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein. All differential PTMs were generated with an expected log fold change of .75 between conditions. Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.

Further details of how the first simulation was generated, can be found in Supplementary Sec. \ref{supp-sec:dataset1}.

@@

In this simulation, data was generated using the same parameters as described in the previous section and including real world experimental conditions. Specifically these simulations included missing values and a low number of modified features. The percentage of missing values and feature counts were determined by looking at the biological experiments in this paper. PTMs were simulated with 2 modified peptide features, while unmodified proteins were still simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. Adding missing values and few representative modified features provides a more realistic expectation of model performance.

Further details of how this simulation was generated, including exact parameters used, can be found in Supplementary Sec. \ref{supp-sec:dataset2}.

\subsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
\label{sec:exp_proc_dataset3}
We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth. This experiment had 4 conditions with two biological replicates per condition and a balanced design.

Experimental Design: In this experiment 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.  Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. Background E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The background lysate was treated as controls and are not expected to be differential in any comparison. The spike-in peptides were mixed with human lysate to create four mixture conditions. Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only.

Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides. There was a 90.2\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

Pairwise Comparisons: We evaluated the ability of the methods to detect known changes in abundance between pairs of conditions. The four mixtures were treated as individual conditions and labeled mix1, mix2, mix3, and mix4. The pairwise comparisons were mix1-mix2, mix1-mix3, mix1-mix4, mix2-mix3, mix2-mix4, and mix3-mix4. The true log fold changes of spike-in modified peptides in these comparisons can be seen in Table \ref{table:spikein_fold_change}.

@@

More information on the experimental design of the benchmark experiment and how the expected fold change was calculated can be found in Supplementary Sec. \ref{supp-sec:benchmark}.

\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
\label{sec:exp_proc_dataset4}
The proposed approach was evaluated on an experiment with six conditions and an unbalanced design.

Experimental Design: In this study Luchetti et al. \cite{LUCHETTI2021} used multiplex proteomics to quantify the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions.

Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides. There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

Pairwise comparisons: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr. All conditions were compared with each other in a full pairwise comparison, resulting in 15 comparisons. Because the dataset is a biological investigation, the true positive modifications were unknown.

Further details on the design of this experiment are provided in Supplementary Sec. \ref{supp-sec:ipah}.

\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
\label{sec:exp_proc_dataset5}
In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced design.

Experimental Design: Maculins et al. \cite{Maculins} designed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri). Multiplex proteomics was used to quantify the abundance of total protein, and phosphorylation in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours). 11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with liquid-chromatography and tandem mass spectrometry (LC-MS/MS) was used. Six conditions were split between 11 channels leading to the experimental design being unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT.

Global Profiling Run: This experiment included a separate global profiling run to measure unmodified peptides. There was a 90\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

Time series comparison: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled KO\_Uninfect, KO\_Early, KO\_Late, WT\_Uninfect, WT\_Early, and WT\_Late. 9 total comparisons were made named KO\_Early-WT\_Early, KO\_Late-WT\_Late, KO\_Uninfected-WT\_Uninfected, KO\_Early-KO\_Uninfected, KO\_Late-KO\_Uninfected, WT\_Early-WT\_Uninfected, WT\_Late-WT\_Uninfected, Infected-Uninfected, and KO-WT. Because the dataset is a biological investigation, the true positive modifications were unknown.

Further details on the design of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.

@@

In this experiment 8 biological replicates were split between 4 conditions in a balanced design.

Experimental Design: Cunningham et al. \cite{Cunningham2015} investigated the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides.

Global Profiling Run: This experiment did not included a separate global profiling run to measure unmodified peptides. In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this leads to substantially less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides. There was a 41.9\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

Pairwise comparisons: We evaluated the statistical approach by its ability to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The four condition were labeled CCCP, Combo, Ctrl, and USP30\_OE. All conditions were compared with each other in a full pairwise comparison, resulting in 6 comparisons. Because the dataset is a biological investigation, the true positive modifications were unknown.

>>>>

\todo{This section would benefit from more biology/technology background, and more extensive references}
Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of bimolecular events. Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen}. Localizing the modification sites, or characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}. For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}. Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale \todo{would be good to have a reference}.

Bottom-up liquid chromatography coupled with tandem mass spectrometry (LC-MS/MS) is a tool of choice for unbiased and large-scale identification and quantification of proteins and their PTMs~\cite{Kall:2011ub,Roepstorff}. However, LC-MS-based interrogation of the modified proteome is challenging, for a number of reasons. First, the relatively lower abundance of modified proteoforms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads \cite{Huang:2014}. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities. Second, contrary to the often large number of identified peptides that can be used to quantify protein abundance, there are relatively few representative peptides that span a modification site, and there may be multiple modified sites on a single peptide~\cite{Mann}. Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complicating the interpretation of the results \cite{Olsen:2013}. Finally, technological aspects of bottom-up MS experiments, such as presence of labeling by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.

The technological challenges in PTM identification and quantification challenge the downstream statistical analyses. Yet frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. In particular, methods such as t-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~ \cite{Ritchie_15a}. Such methods do not fully account for all the sources of uncertainty and variation in the experiment. As the result, they are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as $MSstats$ and $MSstatsTMT$ \cite{Choi:2014,Huang:2020}.

@@

The datasets in this manuscript  benchmarked the proposed approach in situations with known ground truth, and represented a variety of experimental designs and acquisition methods. Two computer simulations varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, PTM types, experimental designs and acquisition strategies. Table \ref{tab:dataDescription} summarizes the experiments. Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and the results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}. Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).

%%%% Summary of controlled mixtures
\begin{table}[h!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of Bio. &  Data \\
&  & Design & Conditions & Replicates & availability \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset in this manuscript.
``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository.  \todo{All the experiments were conducted in data-dependent acquisition (DDA) mode.}
}
\label{tab:dataDescription}
\end{table}

@@

The simulation provides an idealistic case. \todo{generated (how many?)  repeated synthetic datasets?}. \todo{I have a hard time following this section. Would it be possible to rewrite the description, by explaining (1) how many repeated datasets, (2) how many proteins per dataset; (3) how many PTMs per protein, (4) how many modified/unmodified features per protein? Then talk about the quantitative aspects: systematic fold change; random variation; where the parameters of the simulation were inspired from. It will help relate this simulation to real data} had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs. 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions. Of the 500 differential PTMs, the fold change of 250 were fully masked by changes in the unmodified protein. Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein. All differential PTMs were generated with an expected log fold change of .75 between conditions. Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.
Further details can be found in Supplementary Sec. \ref{supp-sec:dataset1}.

@@

The data were simulated as above, while including missing values and a low number of modified features to provide a more realistic representation of the experiments. In this simulation, the feature counts and the proportion of missing values were as observed \todo{on average over all the experimental datasets in this manuscript?}. PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. \todo{There will likely be a comment that missing values are more common among low-abundance features. Did the simulation account for that?} Further details can be found in Supplementary Sec. \ref{supp-sec:dataset2}.

\subsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
\label{sec:exp_proc_dataset3}

The custom spike-in experiment had 4 conditions with two biological replicates per condition, and a balanced design. Additional information can be found in Supplementary Sec. \ref{supp-sec:benchmark}.

\medskip \noindent {\bf Experimental Design:} 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.  Unmodified peptides from Human lysate were used to estimate changes in global protein abundance. Background {\it E. coli} lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The background lysate was treated as control, i.e. it had no expected changes in any comparison.

The spike-in peptides were mixed with human lysate to create four mixture conditions. The true log fold changes of spike-in modified peptides in these comparisons can be seen in Table \ref{table:spikein_fold_change}.  Two sets of data were acquired for each mixture in label-free LC-MS/MS mode: KGG enriched + LC-MS/MS, and LC-MS/MS only. \todo{How the biological replicates in each condition were obtained?}

\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run to measure unmodified peptides. There was a 90.2\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Pairwise Comparisons:} We evaluated the ability of the statistical methods to detect known changes in abundance between pairs of conditions. The four mixtures were treated as individual conditions and labeled mix1, mix2, mix3, and mix4. The pairwise comparisons were mix1-mix2, mix1-mix3, mix1-mix4, mix2-mix3, mix2-mix4, and mix3-mix4.

@@

\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
\label{sec:exp_proc_dataset4}
\todo{In this experiment XXX biological replicates were allocated to 1 TMT mixture in an unbalanced design.} Further details are in Supplementary Sec. \ref{supp-sec:ipah}.

\medskip \noindent {\bf Experimental Design:} Luchetti et al. \cite{LUCHETTI2021} quantified the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The proteome was quantified in 11-plex \todo{?} isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS. \todo{One MS run total?}

\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run \todo{with the same design?} to measure unmodified peptides. There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Pairwise comparisons:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr. All conditions were compared with each other, resulting in 15 pairwise comparisons. Since the dataset was a biological investigation, the true positive modifications were unknown.

\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
\label{sec:exp_proc_dataset5}
In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.

\medskip \noindent {\bf Experimental Design:} Maculins et al. \cite{Maculins} studied primary murine macrophages infected with {\it Shigella flexneri} ({\it S. flexneri}). The experiment quantified the abundance of total protein and of phosphorylation in wild type (WT), and in ATG16L1-deficient (cKO) samples, uninfected and uninfected with {\it S. flexneri}. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).

The proteome was quantified in 11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS. The six conditions were split between 11 channels, and as the result the experiment was unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT.

\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run \todo{with the same design?} to measure unmodified peptides. There was a 90\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Time series comparison:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled KO\_Uninfect, KO\_Early, KO\_Late, WT\_Uninfect, WT\_Early, and WT\_Late. 9 total comparisons were made, namely KO\_Early-WT\_Early, KO\_Late-WT\_Late, KO\_Uninfected-WT\_Uninfected, KO\_Early-KO\_Uninfected, KO\_Late-KO\_Uninfected, WT\_Early-WT\_Uninfected, WT\_Late-WT\_Uninfected, Infected-Uninfected, and KO-WT. Since the dataset was a biological investigation, the true positive modifications were unknown.

@@

In this experiment 8 biological replicates were split between 4 conditions in a balanced design. \todo{Further details in Suppler Sec XXX?}

\medskip \noindent {\bf Experimental Design:} Cunningham et al. \cite{Cunningham2015} investigated the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. The experiment profiled ubiquitination sites, and analyzed changes in the modified site abundance. The experiment had four conditions, CCCP, USP30 over expression (USP30 OE), Combo, and Control. Each condition had two biological replicates. The abundance of modified peptides was quantified with label-free LC-MS/MS.

\medskip \noindent {\bf Global Profiling Run:} This experiment did not include a separate global profiling run to measure unmodified peptides. In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run. However, in addition to low feature counts for unmodified peptides,  this leads to substantially fewer matches between modified and unmodified peptides. There was a 41.9\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Pairwise comparisons:} We evaluated the ability of the statistical methods to detect  changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. All the conditions were compared with each other in a full pairwise comparison, resulting in 6 comparisons. Since the dataset is a biological investigation, the true positive modifications were unknown.
